---
title: "brapiR2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brapiR2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
con <- brapiR2::brapi_connection("https://test-server.brapi.org")
server_up <- isTRUE(tryCatch(
  brapiR2::brapi_ping(con),
  error = function(e) FALSE
))
```

```{r server-down-notice, eval = !server_up, echo = FALSE, results = "asis"}
cat(
  "> **Note:** the public BrAPI test server",
  "(`https://test-server.brapi.org`) was unreachable when this vignette",
  "was built, so the live output below was skipped. The code and its",
  "expected shape are still shown."
)
```

**brapiR2** provides pipe-friendly, stateless access to BrAPI v2 servers.
Unlike stateful clients where you navigate by setting a "current" program or
trial, brapiR2 functions are independent — you pass a connection object and
get a tibble back. This makes them ideal for scripting, pipelines, and
building downstream tools.

## Installation

If you don't already have **remotes** installed, run
`install.packages("remotes")` first.

```{r install, eval = FALSE}
# Not run here: this would reinstall the package while building its own
# documentation.
remotes::install_github("josh45-source/brapiR2")
```

## Connecting to a Server

The examples below run live against the public BrAPI test server, which
requires no authentication.

```{r connect, eval = server_up}
library(brapiR2)

con <- brapi_connection("https://test-server.brapi.org")
con
```

Every function takes `con` as its first argument. No global state is
modified.

## Exploring Programs and Trials

BrAPI organises breeding data as a nesting of four levels, and the calls
below walk down it. A **program** is a long-running breeding effort with a
continuing objective, such as a maize program developing mid-altitude
tropical hybrids; it persists across decades. A **trial** groups experiments
that share a purpose within that program, typically one season's worth of
work, such as the 2022 advanced yield trials. A **study** is a single trial
grown at one location in one season, which is the level at which plots
physically exist and measurements are actually taken. Below that sit
**observation units**, the individual plots, plants or samples that carry the
measurements themselves.

You narrow from one level to the next because each level's identifier is the
filter for the one below it. In practice you rarely know a `studyDbId` up
front; you know you want last season's yield trials from a particular
program, and you find your way down. Note that what a breeder calls an
"environment" in a multi-environment analysis corresponds to a BrAPI study,
not a location, since the same site in two seasons is two environments.

```{r explore, eval = server_up}
library(dplyr)

# List all breeding programs
programs <- brapi_programs(con)
programs

# List trials in the first program
trials <- brapi_trials(con, programDbId = programs$programDbId[1])
trials

# List studies within that trial
studies <- brapi_studies(con, trialDbId = trials$trialDbId[1])
studies
```

## Fetching Phenotypic Data

The `/observations` endpoint returns data in long format, one row per
measurement, with the trait identified in a column and the value in another.
That is the right shape for transport and the wrong shape for analysis:
nearly every R modelling function expects one row per experimental unit and
one column per variable.

`brapi_study_data()` does that reshaping for you. It fetches the study's
observations, pivots traits into columns, and joins the germplasm and plot
identifiers you need to link phenotypes back to genotypes or to a field
layout. Doing this by hand is not difficult, but it is fiddly in ways that
are easy to get subtly wrong, particularly when an observation unit carries
more than one value for the same trait. Repeated measurements are kept as
list-columns rather than silently averaged or dropped, which is why the
summary below unnests before taking means. The function also falls back to
fetching all observations and filtering client-side when a server does not
honour the `studyDbId` filter, so it returns data on servers where a direct
`brapi_observations()` call would come back empty.

```{r pheno, eval = server_up}
# Get analysis-ready wide format: one row per plot, one column per trait
data <- brapi_study_data(con, studies$studyDbId[1])
data
```

The public test server's studies currently have no recorded observations, so
`brapi_study_data()` returns an empty tibble and a warning rather than
fabricated numbers — this is the function's real, documented behavior when
a study has no data, not a bug in the example. Downstream code should check
for this:

```{r pheno-summary, eval = server_up}
if (nrow(data) > 0) {
  trait_cols <- setdiff(
    names(data),
    c(
      "observationUnitDbId", "observationUnitName",
      "germplasmDbId", "germplasmName", "studyDbId", "studyName"
    )
  )
  # Some observation units have more than one recorded value for the same
  # trait (repeated measurements), which brapi_study_data() keeps as a
  # list-column. Unnest those into one row per observation before
  # summarising, so mean() sees plain numbers either way.
  data |>
    tidyr::unnest_longer(dplyr::any_of(trait_cols)) |>
    mutate(across(all_of(trait_cols), as.numeric)) |>
    summarise(across(all_of(trait_cols), \(x) mean(x, na.rm = TRUE)))
} else {
  cat("No observations available for this study on the public test server.\n")
}
```

## Fetching Genotypic Data

A **variant set** is a genotyping dataset: a defined panel of markers scored
on a defined set of samples, usually corresponding to one genotyping run or
one platform. A server may hold several, and they will differ in marker
density, in which individuals were scored, and often in reference genome
build, so picking the right one matters.

The dosage matrix is the form nearly every downstream analysis wants. Each
cell counts how many copies of the alternate allele an individual carries at
that marker, so for diploid maize the values are 0 for homozygous reference,
1 for heterozygous, 2 for homozygous alternate, and `NA` where the call
failed. For polyploids the range extends to the ploidy level, which is why
the function returns counts rather than a genotype string. Representing
genotypes numerically this way is what allows them to enter a mixed model as
a design matrix: additive marker effects are then simply regression
coefficients on allele count.

```{r geno, eval = server_up}
# List available variant sets (genotyping datasets)
vsets <- brapi_variant_sets(con)
vsets

vs_id <- vsets$variantSetDbId[1]
```

`brapi_get_marker_map()` used to read a marker's chromosome and position off
`brapi_variants()`'s `referenceName`/`start` columns. On this server those
come back `NA` for every variant — and per the BrAPI spec, that is legal:
`Variant.start` and `Variant.referenceName` place a variant on a reference
*assembly*, and the spec explicitly allows a server to leave them unset if
it has no such assembly configured for that variant, rather than treating
`NA` as an error condition.

This server has simply chosen not to populate assembly coordinates on its
variants. Positions for the same markers do exist, just under a different
resource: the **Genome Maps** entity (`/maps`, `/markerpositions`) places a
marker on a named *map* instead of a reference assembly — genetic
(measured in cM) or physical (measured in bp), per the map's own `type`
and `unit` fields — and the same marker can appear on more than one map.
`brapi_get_marker_map()` now reads from there instead:

```{r geno-map, eval = server_up}
# Positions for every variant in the set, wherever they have been placed
markers <- brapi_get_marker_map(con, variantSetDbId = vs_id)
markers
```

Not every variant in a set is guaranteed to have a map placement. On this
server, only some of the variant set's variants have one — the warning
above states the exact count rather than silently returning a shorter
tibble. If you already know which map you care about, querying it directly
skips the variant-set lookup and returns the same shape:

```{r geno-map-by-id, eval = server_up}
maps <- brapi_maps(con)
maps[, c("mapDbId", "mapName", "type", "unit")]
```

Both maps on this test server declare a "Physical Map" type with a "cM"
unit. That's inconsistent fixture data on this particular server, not
something brapiR2 or the BrAPI specification produces. On a real server
the two fields agree.

```{r geno-map-by-id-2, eval = server_up}
brapi_get_marker_map(con, mapDbId = maps$mapDbId[1])
```

```{r geno-dosage, eval = server_up}
# Get dosage matrix for genomic selection (samples x markers, values 0/1/2)
dosage <- brapi_get_dosage_matrix(con, vs_id)
dim(dosage)
dosage[seq_len(min(3, nrow(dosage))), seq_len(min(5, ncol(dosage)))]
```

## Authentication

Most production BrAPI servers require authentication. These calls need your
own server and credentials, so they are shown but not run here.

```{r auth, eval = FALSE}
# Username/password login - used by BreedBase, BMS, and Germinate
con <- brapi_connection("https://my-breedbase.org")
con <- brapi_login(con, "username", "password")

# OAuth 2.0 (EBS and similar)
con <- brapi_login_oauth2(
  con,
  client_id     = "my_id",
  client_secret = "my_secret",
  authorize_url = "https://auth.example.org/authorize",
  access_url    = "https://auth.example.org/token"
)

# Bearer token (GIGWA, custom servers)
con <- brapi_set_token(con, Sys.getenv("BRAPI_TOKEN"))
```

### Handling Credentials Safely {#credential-handling}

The example above reads a token from an environment variable rather than
writing it as a literal string, and that's deliberate: a password or token
typed directly into a script gets committed to version control the moment
someone runs `git add`, and stays in the project's history even after it's
"removed" in a later commit. There are two safer places to keep it.

**`.Renviron` and `Sys.getenv()`.** For scripts you run yourself, put
credentials in a `.Renviron` file (in your home directory for all
projects, or in the project directory for that project only) as
`NAME=value` pairs, one per line:

```
BRAPI_TOKEN=abc123...
BRAPI_PASSWORD=my_password
```

R reads `.Renviron` automatically at startup, so after restarting your R
session the values are available via `Sys.getenv()`:

```{r renviron, eval = FALSE}
con <- brapi_connection("https://my-breedbase.org")
con <- brapi_login(
  con, Sys.getenv("BRAPI_USERNAME"), Sys.getenv("BRAPI_PASSWORD")
)
```

Add `.Renviron` to `.gitignore` so it never gets committed. `usethis::edit_r_environ()`
opens the right file for you.

**The keyring package.** For credentials you want stored in your
operating system's own credential manager (Keychain on macOS, Credential
Manager on Windows, Secret Service on Linux) rather than in a plaintext
file, use [keyring](https://cran.r-project.org/package=keyring).
`key_set()` prompts interactively for a secret and stores it; `key_get()`
retrieves it later, in this session or a future one, without the value
ever appearing in your script or your R history:

```{r keyring, eval = FALSE}
# Run once, interactively - prompts for the password and stores it
keyring::key_set("brapiR2_my-breedbase", username = "my_username")

# In scripts, from then on:
con <- brapi_connection("https://my-breedbase.org")
con <- brapi_login(
  con,
  username = "my_username",
  password = keyring::key_get("brapiR2_my-breedbase", username = "my_username")
)
```

Either approach keeps the secret out of the script itself, out of shell
history, and out of anything you might `git commit` or share.

## Caching and Parallel Fetching

These run against the public test server again (caching and parallel
fetching don't need authentication).

```{r performance, eval = server_up}
# Enable caching — repeated calls within the TTL return instantly
cache_dir <- tempfile("brapi_cache_")
dir.create(cache_dir)
perf_con <- brapi_cache_enable(con, ttl = 3600, dir = cache_dir)

# First call: hits the server
invisible(brapi_programs(perf_con))

# Second call: reads from disk (sub-millisecond)
brapi_programs(perf_con)

# Clear all cached files
brapi_cache_clear(perf_con)
```

`brapi_fetch_parallel()` does not set a parallel backend itself - it runs
against whatever `future` plan is already active, sequential by default.
The future package's best-practices vignette is explicit that the choice
of backend belongs to the caller, not to a package: a function that quietly
sets and restores a plan on every call still mutates session-wide state you
didn't ask it to touch, and can silently replace a plan you configured
deliberately (a specific worker count, a cluster spanning several
machines, callr workers for extra isolation). So it's your call to make,
before this function runs:

```{r parallel, eval = server_up}
future::plan(future::multisession, workers = 2)

# Fetch study data from multiple studies in parallel (all studies on the
# server, not just the one attached to the trial filtered above)
study_ids <- brapi_studies(con)$studyDbId
all_data <- brapi_fetch_parallel(perf_con, brapi_study_data, study_ids)
all_data
```

As above, this server's studies have no observations, so `all_data` is an
empty tibble here too — the call itself, and the parallel-fetch mechanics,
are real.

Since you set the plan, shutting the workers back down when you're done
with them is your responsibility too, not something the package does for
you on exit:

```{r cleanup-parallel, eval = server_up}
future::plan(future::sequential)
```

---

## Comparison with QBMS

[QBMS](https://cran.r-project.org/package=QBMS) is a well-established R
package for BrAPI access that has been widely adopted in the breeding
community. brapiR2 and QBMS are **complementary tools designed for
different workflows**.

| Feature | QBMS | brapiR2 |
|---------|------|---------|
| Style | Stateful (set program → set trial → set study) | Stateless (pass `con` everywhere) |
| Returns | Data frames | Tibbles (tidy, list-columns for nested data) |
| Pipe-friendly | Partial | Yes, all functions follow `f(con, ...)` |
| Caching | No | Yes, disk-based with TTL |
| Parallel fetch | No | Via the caller's own `future` backend |
| Genotypics | Limited | Allele matrix, dosage matrix, marker map (32/36 BrAPI entities covered overall, read-only) |
| Authentication | Yes | Yes |

brapiR2 targets BrAPI v2 only, by design - see the "Design History" section
of [DESIGN.md](https://github.com/josh45-source/brapiR2/blob/main/DESIGN.md)
for why. QBMS's own BrAPI version support is a separate question from this
comparison; check QBMS's own documentation for its current v1/v2 coverage
rather than relying on a claim made here.

### Same workflow, different style

The examples below use a private BreedBase-style server and credentials, so
they are illustrative only.

**QBMS** — navigational, stateful:

```{r qbms-example, eval = FALSE}
library(QBMS)

set_crop("Wheat")
login_bms("https://my-bms.org", "user", "pass")

list_programs()
set_program("Wheat Breeding")

list_trials()
set_trial("Yield Trial 2022")

list_studies()
set_study("Ithaca 2022")

data <- get_study_data()
```

**brapiR2** — functional, pipe-friendly:

```{r brapiR2-example, eval = FALSE}
library(brapiR2)
library(dplyr)

con <- brapi_connection("https://my-bms.org") |>
  brapi_login("user", "pass")

data <- brapi_programs(con) |>
  filter(programName == "Wheat Breeding") |>
  pull(programDbId) |>
  (\(pid) brapi_trials(con, programDbId = pid))() |>
  filter(trialName == "Yield Trial 2022") |>
  pull(trialDbId) |>
  (\(tid) brapi_studies(con, trialDbId = tid))() |>
  filter(studyName == "Ithaca 2022") |>
  pull(studyDbId) |>
  brapi_study_data(con = con)
```

## References

- Selby, P., Abbeloos, R., Backlund, J.E., et al., & The BrAPI Consortium
  (2019). BrAPI — an application programming interface for plant breeding
  applications. *Bioinformatics*, 35(20), 4147–4155.
  <https://doi.org/10.1093/bioinformatics/btz190>

Choose QBMS for interactive exploration in the console. Choose brapiR2 for
reproducible scripts, pipelines, or when you need genotypic data and
caching. Many users use both.
