---
title: "Genomic Selection Pipeline with brapiR2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genomic Selection Pipeline with brapiR2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates a complete genomic selection (GS) workflow using
brapiR2 to pull phenotypic and genotypic data directly from a BrAPI-compliant
server into R, then running GS models with rrBLUP.

The `brapi_get_dosage_matrix()` function returns a standard numeric matrix
(samples × markers) that is directly compatible with the major GS packages:

| Package | Usage |
|---------|-------|
| [rrBLUP](https://cran.r-project.org/package=rrBLUP) | `mixed.solve(y, Z = dosage)` |
| [BGLR](https://cran.r-project.org/package=BGLR) | `ETA = list(MRK = list(X = dosage, model = "BRR"))` |
| [sommer](https://cran.r-project.org/package=sommer) | `mmer(yield ~ 1, random = ~ vsr(germplasmName, Gu = A))` |
| [AGHmatrix](https://cran.r-project.org/package=AGHmatrix) | `Gmatrix(dosage, method = "VanRaden")` |

## Step 1: Connect and Fetch Phenotypic Data

```{r pheno}
library(brapiR2)
library(dplyr)

con <- brapi_connection("https://my-breedbase.org")
con <- brapi_login(con, "username", "password")

# Optional: enable caching so re-runs don't hit the server again
con <- brapi_cache_enable(con, ttl = 7200)

# Get phenotype data in wide format (one row per plot, one column per trait)
pheno_raw <- brapi_study_data(con, "my_study_id")
pheno_raw
#> # A tibble: 250 x 8
#>   observationUnitDbId germplasmDbId germplasmName  studyDbId
#>   <chr>               <chr>         <chr>          <chr>
#> 1 ou001               g001          Variety_001    study_01
#> 2 ou002               g002          Variety_002    study_01
#> # ... with 246 more rows, and 4 more variables:
#> #   Grain_Yield_t_ha <chr>, Days_to_Heading <chr>,
#> #   Plant_Height_cm <chr>, Thousand_Grain_Weight <chr>

# Convert trait columns to numeric and drop missing
pheno <- pheno_raw |>
  mutate(across(Grain_Yield_t_ha:Thousand_Grain_Weight, as.numeric)) |>
  filter(!is.na(Grain_Yield_t_ha))

nrow(pheno)
#> [1] 241
```

## Step 2: Fetch Genotypic Data

```{r geno}
# List available variant sets to find the right genotyping dataset
vsets <- brapi_variant_sets(con)
vsets
#> # A tibble: 3 x 8
#>   variantSetDbId variantSetName callSetCount variantCount
#>   <chr>          <chr>                 <int>        <int>
#> 1 vs001          SNP50K_2022              250        45230
#> 2 vs002          GBS_v2_2022              250        82100
#> 3 vs003          DArTseq_panel            250        14500

# Get dosage matrix: rows = samples (callSetDbIds), cols = markers (variantDbIds)
# Values: 0 (hom ref), 1 (het), 2 (hom alt), NA (missing)
dosage <- brapi_get_dosage_matrix(con, "vs001")
#> i Fetching allele matrix for variant set 'vs001'...
#> i Encoding 11307500 genotype calls as allele dosages...
#> v Dosage matrix ready: 250 samples x 45230 markers.

dim(dosage)
#> [1]   250 45230

# Preview dosage values
dosage[1:4, 1:6]
#>             SNP_chr1_1234 SNP_chr1_5678 SNP_chr2_234 SNP_chr2_890 ...
#> Variety_001             0             1            2            0
#> Variety_002             2             0            0            1
#> Variety_003             1             1            1            1
#> Variety_004             0             0            2            0

# Get marker map for downstream QTL / Manhattan plots
markers <- brapi_get_marker_map(con, "vs001")
markers
#> # A tibble: 45230 x 4
#>    variantDbId    variantName   referenceName  start
#>    <chr>          <chr>         <chr>          <int>
#>  1 var001         SNP_chr1_1234 1A              1234
#>  2 var002         SNP_chr1_5678 1A              5678
#>  # ... with 45228 more rows
```

## Step 3: Align Phenotypes and Genotypes

```{r align}
# Match on germplasmName (rownames of dosage = callSetDbId = germplasmName
# on most servers; adjust if your server uses a different identifier)
common <- intersect(rownames(dosage), pheno$germplasmName)
length(common)
#> [1] 237

pheno_aligned <- pheno |>
  filter(germplasmName %in% common) |>
  arrange(match(germplasmName, common))

geno_aligned <- dosage[common, ]

# Quick missing-data summary
mean(is.na(geno_aligned))
#> [1] 0.0312     # ~3.1% missing calls — impute before modelling
```

## Step 4: Imputation (Optional but Recommended)

```{r impute}
# Simple mean imputation (column means)
col_means <- colMeans(geno_aligned, na.rm = TRUE)
for (j in seq_len(ncol(geno_aligned))) {
  geno_aligned[is.na(geno_aligned[, j]), j] <- col_means[j]
}

# Or use rrBLUP's A.mat which handles missing data internally
```

## Step 5: Run Genomic Selection with rrBLUP

```{r gs-rrBLUP}
library(rrBLUP)

# Mixed model: phenotype ~ overall mean + random marker effects
result <- mixed.solve(
  y = pheno_aligned$Grain_Yield_t_ha,
  Z = geno_aligned
)

# Estimated marker effects
length(result$u)
#> [1] 45230

# GEBVs (Genomic Estimated Breeding Values)
gebvs <- geno_aligned %*% result$u + result$beta
head(gebvs)
#>               [,1]
#> Variety_001  4.82
#> Variety_002  5.14
#> Variety_003  4.67
#> Variety_004  5.31
#> Variety_005  4.91
#> Variety_006  5.09

hist(gebvs,
  main = "Distribution of GEBVs — Grain Yield",
  xlab = "GEBV (t/ha)", col = "steelblue"
)
```

## Step 6: Alternative GS Packages

```{r gs-other}
# ---- BGLR (Bayesian regression) ----
library(BGLR)

fm_bglr <- BGLR(
  y = pheno_aligned$Grain_Yield_t_ha,
  ETA = list(MRK = list(X = geno_aligned, model = "BRR")),
  nIter = 6000, burnIn = 1000,
  verbose = FALSE
)
gebvs_bglr <- geno_aligned %*% fm_bglr$ETA$MRK$b

# ---- sommer (mixed model with G matrix) ----
library(sommer)
library(AGHmatrix)

# Build genomic relationship matrix
G <- Gmatrix(geno_aligned, method = "VanRaden")
#> Gmatrix computation: VanRaden method
#> Markers: 45230  Individuals: 237

fm_sommer <- mmer(
  Grain_Yield_t_ha ~ 1,
  random  = ~ vsr(germplasmName, Gu = G),
  rcov    = ~units,
  data    = pheno_aligned
)
gebvs_sommer <- randef(fm_sommer)$`u:germplasmName`
```

## Step 7: Cross-Validation

```{r cv}
set.seed(42)
n <- nrow(pheno_aligned)
folds <- sample(rep(1:5, length.out = n))
cors <- numeric(5)

for (k in 1:5) {
  train <- folds != k
  test <- folds == k

  fit <- mixed.solve(
    y = pheno_aligned$Grain_Yield_t_ha[train],
    Z = geno_aligned[train, ]
  )

  pred <- geno_aligned[test, ] %*% fit$u + fit$beta
  cors[k] <- cor(as.numeric(pred), pheno_aligned$Grain_Yield_t_ha[test])
}

cat(sprintf(
  "5-fold CV prediction accuracy: %.3f ± %.3f\n",
  mean(cors), sd(cors)
))
#> 5-fold CV prediction accuracy: 0.612 ± 0.054
```
