---
title: "Multi-Environment Trial (MET) Analysis with brapiR2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Environment Trial (MET) Analysis with brapiR2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Multi-environment trial (MET) analysis is the core workflow of a plant breeding
program: gather phenotypic observations from many locations and years, then
decompose genotype × environment (GxE) interaction to identify stable,
high-performing varieties.

This vignette shows how to:

1. Pull phenotypic data from multiple BrAPI studies in parallel
2. Attach environment metadata (location, season)
3. Run AMMI / GGE biplots with the **metan** package
4. Run mixed-model stability analysis with **lme4**

## Step 1: Connect and Discover Studies

```{r setup}
library(brapiR2)
library(dplyr)

con <- brapi_connection("https://my-breedbase.org")
con <- brapi_login(con, "username", "password")

# Cache responses to speed up re-runs and protect server bandwidth
con <- brapi_cache_enable(con, ttl = 7200)
#> v Caching enabled at ~/.cache/brapiR2 (TTL: 7200s)
```

```{r discover}
# Discover all studies in the target trial
trials <- brapi_trials(con, programDbId = "wheat_program_01")
trials
#> # A tibble: 4 x 12
#>   trialDbId   trialName           programDbId active
#>   <chr>       <chr>               <chr>       <lgl>
#> 1 trial_2022  National Yield 2022 wheat_prog  TRUE
#> 2 trial_2021  National Yield 2021 wheat_prog  FALSE
#> 3 trial_2020  National Yield 2020 wheat_prog  FALSE
#> 4 trial_drght Drought Nursery     wheat_prog  TRUE

# Focus on the current MET
studies <- brapi_studies(con, trialDbId = "trial_2022")
studies
#> # A tibble: 8 x 15
#>   studyDbId  studyName          locationDbId locationName  seasonDbId
#>   <chr>      <chr>              <chr>        <chr>         <chr>
#> 1 study_e1   Ithaca NY 2022     loc_ithaca   Ithaca        season_2022
#> 2 study_e2   Geneva NY 2022     loc_geneva   Geneva        season_2022
#> 3 study_e3   Riverhead NY 2022  loc_river    Riverhead     season_2022
#> 4 study_e4   State College 2022 loc_psu      State College season_2022
#> 5 study_e5   Wooster OH 2022    loc_wooster  Wooster       season_2022
#> 6 study_e6   Lexington KY 2022  loc_lexing   Lexington     season_2022
#> 7 study_e7   Raleigh NC 2022    loc_raleigh  Raleigh       season_2022
#> 8 study_e8   Gainesville FL 22  loc_gnsvlle  Gainesville   season_2022

study_ids <- studies$studyDbId
```

## Step 2: Fetch All Study Data in Parallel

`brapi_fetch_parallel()` uses the **furrr** package to run one worker per study
concurrently. With 8 environments and 4 workers, the fetch completes in roughly
the time of 2 sequential requests.

```{r parallel-fetch}
library(future) # plan() sets the parallel backend

# Adjust .workers to match your CPU count or server rate limits
met_raw <- brapi_fetch_parallel(
  con, brapi_study_data, study_ids,
  .workers = 4L
)
#> i Fetching 8 items across 4 workers...
#> v Fetched 2400 rows from 8 sources.

glimpse(met_raw)
#> Rows: 2,400
#> Columns: 10
#> $ observationUnitDbId  <chr> "ou_e1_001", "ou_e1_002", ...
#> $ observationUnitName  <chr> "Plot 101", "Plot 102", ...
#> $ germplasmDbId        <chr> "g001", "g002", ...
#> $ germplasmName        <chr> "Variety_001", "Variety_002", ...
#> $ studyDbId            <chr> "study_e1", "study_e1", ...
#> $ studyName            <chr> "Ithaca NY 2022", ...
#> $ Grain_Yield_t_ha     <chr> "5.12", "4.89", ...
#> $ Days_to_Heading      <chr> "65", "68", ...
#> $ Plant_Height_cm      <chr> "92", "88", ...
#> $ Thousand_Grain_Wt    <chr> "38.4", "41.2", ...
```

## Step 3: Attach Environment Metadata

The study data already carries `studyDbId` and `studyName`. Join with location
and season data from the `studies` tibble for richer environment labels.

```{r env-metadata}
# Bring in location and season columns from the studies tibble
env_meta <- studies |>
  select(studyDbId, studyName, locationName, seasonDbId)

met <- met_raw |>
  left_join(env_meta, by = "studyDbId") |>
  mutate(
    Grain_Yield_t_ha = as.numeric(Grain_Yield_t_ha),
    Days_to_Heading = as.numeric(Days_to_Heading),
    Plant_Height_cm = as.numeric(Plant_Height_cm),
    Thousand_Grain_Wt = as.numeric(Thousand_Grain_Wt),
    env = locationName # short environment label
  ) |>
  filter(!is.na(Grain_Yield_t_ha))

nrow(met)
#> [1] 2311   # some missing values removed

# Confirm we have data from all 8 environments
table(met$env)
#>   Gainesville       Geneva  Ithaca  Lexington  Raleigh  Riverhead
#>           285          289     291         278      283        286
#>  State College  Wooster
#>            304      295
```

## Step 4: Means Matrix for GxE Analysis

Many GxE methods (AMMI, GGE biplot) require a two-way table of genotype means
per environment.

```{r means-matrix}
library(tidyr)

# Compute adjusted means (plot-level mean per genotype × environment)
means <- met |>
  group_by(germplasmName, env) |>
  summarise(yield = mean(Grain_Yield_t_ha, na.rm = TRUE), .groups = "drop")

# Wide format: rows = genotypes, columns = environments
means_wide <- means |>
  pivot_wider(names_from = env, values_from = yield)

means_wide
#> # A tibble: 300 x 9
#>   germplasmName  Gainesville Geneva Ithaca Lexington Raleigh Riverhead ...
#>   <chr>                <dbl>  <dbl>  <dbl>     <dbl>   <dbl>     <dbl>
#> 1 Variety_001           5.23   4.98   5.12      5.41    4.87      5.03
#> 2 Variety_002           4.89   5.34   4.91      4.72    5.12      5.28
#> # ...
```

## Step 5: AMMI and GGE Biplot with metan

```{r metan-ammi}
library(metan)

# AMMI model: decompose GxE with additive main effects + multiplicative interaction
ammi_model <- performs_ammi(
  .data = met,
  env = env,
  gen = germplasmName,
  rep = observationUnitDbId, # plot as replicate; adjust if you have block IDs
  resp = Grain_Yield_t_ha,
  verbose = FALSE
)

# AMMI stability value (ASV) — lower = more stable across environments
ammi_stable <- ammi_indexes(ammi_model)
ammi_stable$Grain_Yield_t_ha |>
  arrange(ASV) |>
  head(10)
#> # A tibble: 10 x 6
#>   GEN          Y     IPCA1   IPCA2    ASV  WAAS
#>   <chr>    <dbl>     <dbl>   <dbl>  <dbl> <dbl>
#> 1 Var_089  5.34    0.00312  0.0041  0.012  0.51
#> 2 Var_214  5.21    0.00891  0.0023  0.035  0.48
#> # ...

# AMMI1 biplot: IPCA1 scores vs mean yield
plot(ammi_model, type = "AMMI1")
```

```{r metan-gge}
# GGE biplot
gge_model <- gge(
  .data = met,
  env   = env,
  gen   = germplasmName,
  rep   = observationUnitDbId,
  resp  = Grain_Yield_t_ha
)

plot(gge_model, type = "which_won_where")
#> (biplot showing mega-environments and winning genotypes per environment)
```

## Step 6: Mixed Model Stability with lme4

For unbalanced data or when you need BLUPs rather than means, fit a mixed model
with genotype × environment interaction as a random effect.

```{r lme4}
library(lme4)

# Basic mixed model: GxE interaction as random
fm <- lmer(
  Grain_Yield_t_ha ~ env + (1 | germplasmName) + (1 | germplasmName:env),
  data = met
)
summary(fm)
#> Random effects:
#>  Groups               Name        Variance Std.Dev.
#>  germplasmName:env    (Intercept) 0.0812   0.285
#>  germplasmName        (Intercept) 0.1947   0.441
#>  Residual                         0.0534   0.231

# Extract BLUPs for genotypic effects
blups <- ranef(fm)$germplasmName
blups <- tibble(
  germplasmName = rownames(blups),
  BLUP          = blups[["(Intercept)"]]
) |>
  arrange(desc(BLUP))

head(blups, 10)
#> # A tibble: 10 x 2
#>   germplasmName   BLUP
#>   <chr>          <dbl>
#> 1 Variety_089    0.412
#> 2 Variety_214    0.389
#> 3 Variety_031    0.371
#> # ...

# Variance component partitioning
VarCorr(fm)
#>  Groups            Name        Std.Dev.
#>  germplasmName:env (Intercept)  0.285    # GxE component
#>  germplasmName     (Intercept)  0.441    # G (genetic signal)
#>  Residual                       0.231

# H^2 broad-sense heritability (across-environment)
var_g <- as.numeric(VarCorr(fm)$germplasmName)
var_ge <- as.numeric(VarCorr(fm)$`germplasmName:env`)
var_e <- sigma(fm)^2
n_env <- length(unique(met$env))
n_rep <- 1 # one plot per genotype per env (adjust if replicated)

H2 <- var_g / (var_g + var_ge / n_env + var_e / (n_env * n_rep))
cat(sprintf("Broad-sense H^2 (across environments): %.3f\n", H2))
#> Broad-sense H^2 (across environments): 0.731
```

## Step 7: Final Selection List

```{r selection}
# Combine BLUP ranking with stability
selection <- blups |>
  left_join(
    ammi_stable$Grain_Yield_t_ha |>
      select(GEN, mean_yield = Y, ASV) |>
      rename(germplasmName = GEN),
    by = "germplasmName"
  ) |>
  mutate(
    # Normalise both metrics (0 = worst, 1 = best)
    rank_yield = rank(BLUP) / n(),
    rank_stab  = 1 - rank(ASV) / n(), # lower ASV = more stable = better
    score      = 0.6 * rank_yield + 0.4 * rank_stab
  ) |>
  arrange(desc(score))

# Top 20 selections: high yield + high stability
top20 <- head(selection, 20)
top20
#> # A tibble: 20 x 6
#>   germplasmName    BLUP mean_yield   ASV rank_yield rank_stab score
#>   <chr>           <dbl>      <dbl> <dbl>      <dbl>     <dbl> <dbl>
#> 1 Variety_089     0.412       5.34 0.012      0.998     0.997 0.998
#> 2 Variety_214     0.389       5.21 0.035      0.991     0.989 0.991
#> # ...
```
