---
title: "Genomic Selection Pipeline with brapiR2"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
con <- brapiR2::brapi_connection("https://test-server.brapi.org")
server_up <- isTRUE(tryCatch(
  brapiR2::brapi_ping(con),
  error = function(e) FALSE
))
```

```{r server-down-notice, eval = !server_up, echo = FALSE, results = "asis"}
cat(
  "> **Note:** the public BrAPI test server",
  "(`https://test-server.brapi.org`) was unreachable when this article was",
  "built, so Part 1's live output below was skipped. Part 3 (simulated",
  "data) is unaffected and still ran."
)
```

## Overview

This article demonstrates a complete genomic selection (GS) workflow using
brapiR2 to pull phenotypic and genotypic data from a BrAPI-compliant server
into R, then running GS models with rrBLUP, BGLR, sommer, and AGHmatrix.

The `brapi_get_dosage_matrix()` function returns a standard numeric matrix
(samples x markers) that is directly compatible with the major GS packages:

| Package | Usage |
|---------|-------|
| [rrBLUP](https://cran.r-project.org/package=rrBLUP) | `mixed.solve(y, Z = dosage)` |
| [BGLR](https://cran.r-project.org/package=BGLR) | `ETA = list(MRK = list(X = dosage, model = "BRR"))` |
| [sommer](https://cran.r-project.org/package=sommer) | `mmer(yield ~ 1, random = ~ vsr(germplasmName, Gu = A))` |
| [AGHmatrix](https://cran.r-project.org/package=AGHmatrix) | `Gmatrix(dosage, method = "VanRaden")` |

It has three parts: **Part 1** fetches real data from the public BrAPI test
server and shows exactly what brapiR2 returns. **Part 2** shows the
authenticated pattern you'd use against your own server, without running it.
**Part 3** runs the actual GS modeling on a simulated dataset, because (as
Part 1 demonstrates directly) the public server doesn't have enough real
individuals with both phenotype and genotype data to fit and validate a
model.

## Part 1: Live Data from the Public Test Server

These calls run against `https://test-server.brapi.org` and require no
authentication.

```{r pheno-live, eval = server_up}
library(brapiR2)
library(dplyr)

studies <- brapi_studies(con)
studies

# brapi_study_data() falls back to fetching all observations and filtering
# client-side when a server (like this one) doesn't honour the studyDbId
# filter on /observations - so it succeeds even though a direct
# brapi_observations(con, studyDbId = ...) call would come back empty.
pheno_by_study <- lapply(studies$studyDbId, brapi_study_data, con = con)
names(pheno_by_study) <- studies$studyDbId
pheno_by_study[["study1"]]
```

```{r geno-live, eval = server_up}
vsets <- brapi_variant_sets(con)
vsets

vs_id <- vsets$variantSetDbId[1]

## brapi_variants()'s referenceName/start are NA on this server (the BrAPI
## spec makes them optional - they place a variant on a reference
## assembly, and this server has none configured). brapi_get_marker_map()
## instead reads the Genome Maps entity (/markerpositions), which places
## each marker on a named map (genetic, cM, or physical, bp) and is
## populated here.
markers <- brapi_get_marker_map(con, variantSetDbId = vs_id)
markers

dosage <- brapi_get_dosage_matrix(con, vs_id)
dim(dosage)
dosage[seq_len(min(5, nrow(dosage))), seq_len(min(6, ncol(dosage)))]
```

### Do phenotype and genotype records share any germplasm here?

This is the question that decides whether real genomic prediction is
possible against this server. `brapi_call_sets()` links a genotyped sample
to a `sampleDbId`, and `brapi_samples()` links that same `sampleDbId` to a
`germplasmDbId` - joining on `sampleDbId` (not `callSetDbId`; they're
different ID spaces on this server).

```{r overlap-live, eval = server_up}
pheno_germplasm <- unique(unlist(lapply(pheno_by_study, function(d) {
  if (nrow(d) > 0 && "germplasmDbId" %in% names(d)) {
    d$germplasmDbId
  } else {
    character(0)
  }
})))
pheno_germplasm

call_sets <- brapi_call_sets(con, variantSetDbId = vs_id)
samples <- brapi_samples(con)

geno_germplasm <- samples |>
  filter(.data$sampleDbId %in% call_sets$sampleDbId) |>
  pull(.data$germplasmDbId) |>
  unique()
geno_germplasm

overlap <- intersect(pheno_germplasm, geno_germplasm)
overlap
```

Real, live result: `r if (server_up) length(overlap) else "(server unreachable at build time)"`
germplasm have both phenotype and genotype records on this server today.
That is genuinely more than zero, but far short of the dozens of individuals
a real genomic prediction exercise needs to fit a model and hold out a
validation set. See
[`dev/explore-test-server.R`](https://github.com/josh45-source/brapiR2/blob/main/dev/explore-test-server.R)
for the full survey this is drawn from. Part 3 below uses simulated data so
the modeling section can actually run.

## Part 2: Authenticated Access to Your Own Server (illustrative)

Production BrAPI servers (BreedBase, BMS, Germinate, ...) require
credentials the public test server doesn't need. This shows the pattern —
it is not run here, since it needs a real server and login of your own.

```{r pheno-illustrative, eval = FALSE}
con_priv <- brapi_connection("https://my-breedbase.org")
con_priv <- brapi_login(con_priv, "username", "password")

# Optional: enable caching so re-runs don't hit the server again
con_priv <- brapi_cache_enable(con_priv, ttl = 7200)

# Get phenotype data in wide format (one row per plot, one column per trait)
pheno_raw <- brapi_study_data(con_priv, "my_study_id")
pheno_raw

# Convert trait columns to numeric and drop missing
pheno <- pheno_raw |>
  mutate(across(Grain_Yield_t_ha:Thousand_Grain_Weight, as.numeric)) |>
  filter(!is.na(Grain_Yield_t_ha))
```

```{r geno-illustrative, eval = FALSE}
# List available variant sets to find the right genotyping dataset
vsets_priv <- brapi_variant_sets(con_priv)
vsets_priv

# Get dosage matrix: rows = samples (callSetDbIds),
# cols = markers (variantDbIds). Values: 0 (hom ref), 1 (het), 2 (hom alt),
# NA (missing)
dosage_priv <- brapi_get_dosage_matrix(con_priv, "vs001")
dim(dosage_priv)

# Get marker map for downstream QTL / Manhattan plots
markers_priv <- brapi_get_marker_map(con_priv, variantSetDbId = "vs001")
markers_priv
```

## Part 3: Genomic Selection on Simulated Data

Genomic prediction needs the same germplasm scored on both sides — enough of
them to fit marker effects and validate the fit. Part 1 showed real
brapiR2 output; it also showed why the public server can't support the
model-fitting step: only `r if (server_up) length(overlap) else 2` real
germplasm have paired data there, and no public BrAPI server currently
provides the dozens-to-hundreds of paired records genomic prediction
actually needs. The rest of this article uses a **simulated** dataset built
with [AlphaSimR](https://cran.r-project.org/package=AlphaSimR), generated
with a fixed seed so it is fully reproducible without server access.
Simulated values are labeled as such throughout.

```{r simulate-founders, eval = FALSE}
# SIMULATED founder haplotypes. runMacs() runs a coalescent simulation
# (MaCS; Chen, Marjoram & Wall 2009) that produces markers with realistic
# linkage disequilibrium (LD) and shared ancestry between founders - unlike
# the independent per-marker draws used in earlier drafts of this article.
#
# Not run here: AlphaSimR does not expose R-level seed control for this
# coalescent step (set.seed() has no effect on runMacs() itself - verified,
# two runs with the same seed produce different founder haplotypes), so
# calling it at build time would make every number below change on every
# rebuild of this article. Instead, this call was run once, and its result
# saved to founderPop.rds (loaded below) so the article - including the
# reported cross-validation accuracy - is exactly reproducible.
library(AlphaSimR)

n_founders <- 150L # founder lines
n_sites    <- 800L # segregating sites simulated per chromosome

founderPop <- runMacs(
  nInd = n_founders, nChr = 1L, segSites = n_sites,
  species = "GENERIC", nThreads = 1L
)
saveRDS(founderPop, "founderPop.rds")
```

```{r simulate}
library(dplyr)
library(AlphaSimR)

n_founders <- 150L # founder lines (must match the frozen founderPop.rds)
n_qtl      <- 100L # of the founder genome's segregating sites, how many are QTL
n_snps     <- 700L # SNP chip size: the remaining sites, QTL excluded
h2         <- 0.5  # target narrow-sense heritability

# Load the frozen founder population (see the eval = FALSE chunk above for
# how it was made) instead of calling runMacs() here, so this article's
# genotypes - and the cross-validation accuracy reported at the end of this
# section - are exactly reproducible across rebuilds. Every step *after*
# this one (trait assignment, crossing, cross-validation folds) was already
# fully reproducible given the founder population it starts from; freezing
# the founders makes the whole pipeline reproducible end to end.
founderPop <- readRDS("founderPop.rds")

set.seed(42)

# SimParam defines the trait architecture on top of the founder genome:
# n_qtl randomly chosen segregating sites carry additive effects drawn to
# hit the target heritability; the remaining sites are neutral markers
# only informative through their LD with the causal ones. That's now a
# fact about the genome, not yet about what the model sees - the SNP chip
# defined next is drawn separately, only from those neutral sites, so it
# deliberately excludes the QTL themselves.
SP <- SimParam$new(founderPop)
SP$addTraitA(nQtlPerChr = n_qtl, mean = 5, var = 1) # SIMULATED yield (t/ha)
SP$setVarE(h2 = h2)

# A SNP chip is a genotyping platform's fixed marker panel, decided before
# anyone knows which sites are causal. addSnpChip() models that: it draws
# its n_snps markers only from the segregating sites addTraitA() did NOT
# designate as QTL, so every marker on the chip tags the causal variants
# only through LD, never by being one.
SP$addSnpChip(nSnpPerChr = n_snps)

founders <- newPop(founderPop, simParam = SP)

# One round of random crossing: each of n_founders progeny descends from
# two randomly chosen founders. Unlike independent markers, the resulting
# population has real family relatedness and real recombination
# breakpoints between founder haplotypes.
progeny <- randCross(
  founders, nCrosses = n_founders, nProgeny = 1L, simParam = SP
)

# pullSnpGeno() returns the dosage matrix for the SNP chip only, not the
# QTL: rows = individuals, columns = chip markers, values = 0/1/2 copies
# of the alternate allele - the same shape brapi_get_dosage_matrix()
# returns in Part 1. This is the deliberate difference from
# pullSegSiteGeno(), used in earlier drafts of this article: a real
# marker panel never includes the causal variants themselves, so a
# simulation that hands the model the true QTL directly would overstate
# how well genomic prediction works on a real genotyping platform.
geno_aligned <- pullSnpGeno(progeny, simParam = SP)

pheno_aligned <- tibble(
  germplasmName    = progeny@id,
  Grain_Yield_t_ha = pheno(progeny)[, 1] # SIMULATED yield (t/ha)
)

cat(sprintf(
  "Simulated %d lines x %d SNP-chip markers, %d QTL (not on the chip), target h2 = %.2f\n",
  nrow(geno_aligned), ncol(geno_aligned), n_qtl, h2
))
```

As a sanity check, adjacent SNP-chip markers should be far more correlated
with each other than distant ones - that's LD, and it's what genomic
prediction exploits when the causal variants themselves aren't observed.
The independent-marker simulation in earlier drafts of this article had no
such structure by construction.

```{r ld-check}
r2_adjacent <- vapply(seq_len(ncol(geno_aligned) - 1), function(j) {
  suppressWarnings(cor(geno_aligned[, j], geno_aligned[, j + 1])^2)
}, numeric(1))

set.seed(45)
r2_random <- vapply(seq_len(200), function(i) {
  j <- sample.int(ncol(geno_aligned), 1)
  j2 <- sample(setdiff(seq_len(ncol(geno_aligned)), j), 1)
  suppressWarnings(cor(geno_aligned[, j], geno_aligned[, j2])^2)
}, numeric(1))

cat(sprintf(
  "Mean r^2, adjacent markers: %.3f | random marker pairs: %.3f\n",
  mean(r2_adjacent, na.rm = TRUE), mean(r2_random, na.rm = TRUE)
))
```

```{r simulate-missing}
# Introduce a small amount of missing genotype data, as real platforms do.
set.seed(43)
n_missing <- floor(0.03 * length(geno_aligned))
missing_idx <- sample.int(length(geno_aligned), n_missing)
geno_aligned[missing_idx] <- NA

mean(is.na(geno_aligned))
```

### Imputation

Missing calls have to go somewhere before modelling, because the methods
below are matrix operations with no notion of a missing cell. `mixed.solve()`
will not accept `NA` in its design matrix, and dropping every individual with
any missing call would discard most of a real panel, since missingness is
scattered rather than concentrated.

The usual first step is not imputation at all but filtering: markers missing
in more than roughly 10 to 20 percent of individuals, and individuals missing
more than roughly 20 to 30 percent of their calls, are typically removed
rather than reconstructed, because there is too little information left to
reconstruct them from. What remains is imputed. Below about 5 percent
residual missingness, which is where this simulation sits, the choice of
method makes very little difference to prediction accuracy.

Mean imputation, used here, replaces a missing call with the marker's mean
dosage across individuals. It is unbiased for allele frequency and costs
nothing, but it pulls imputed individuals toward the population mean and so
slightly shrinks genetic variance. Rounding to whole numbers, which we do
here because `AGHmatrix::Gmatrix()` expects integer calls, discards the
fractional information mean imputation supplies and is a further small
compromise. The alternative shown in the comment, `rrBLUP::A.mat()`, handles
missingness while computing the relationship matrix and so avoids modifying
the genotype matrix at all. Where imputation genuinely matters, with high
missingness or low marker density, a haplotype-based imputer such as Beagle
that borrows information from linked markers and shared haplotypes will beat
both, at the cost of an external tool and a reference panel.

```{r impute}
# Simple mean imputation (column means), rounded back to whole-number
# dosages since AGHmatrix::Gmatrix() expects integer genotype calls.
col_means <- round(colMeans(geno_aligned, na.rm = TRUE))
for (j in seq_len(ncol(geno_aligned))) {
  geno_aligned[is.na(geno_aligned[, j]), j] <- col_means[j]
}

anyNA(geno_aligned)

# Or use rrBLUP's A.mat, which handles missing data internally.
```

### Run Genomic Selection with rrBLUP

`mixed.solve()` fits `y = mu + Zu + e`, where `Z` is the dosage matrix and
`u` is a vector of marker effects treated as random draws from a common
normal distribution. That last assumption is the substantive one: it says
every marker contributes a small effect drawn from the same distribution,
which is the infinitesimal model. Its practical consequence is that all
effects are shrunk toward zero by an amount the model estimates from the
data, so with far more markers than individuals the fit stays well behaved
instead of chasing noise.

The function returns `result$u`, one estimated effect per marker, and
`result$beta`, the intercept. A genomic estimated breeding value is then just
the sum of an individual's marker effects weighted by its allele dosages,
which is the matrix product computed below. GEBVs come out in trait units,
tonnes per hectare here, and represent predicted genetic merit rather than
expected phenotype: they carry the heritable component only, which is exactly
what you want when ranking candidates for selection. This ridge formulation
is mathematically equivalent to GBLUP fitted through a genomic relationship
matrix; the two differ in computation, not in the model.

```{r gs-rrBLUP}
library(rrBLUP)

# Mixed model: phenotype ~ overall mean + random marker effects
result <- mixed.solve(
  y = pheno_aligned$Grain_Yield_t_ha,
  Z = geno_aligned
)

# Estimated marker effects
length(result$u)

# GEBVs (Genomic Estimated Breeding Values)
# mixed.solve() returns u/beta as 1-D arrays; as.vector() drops the dim
# attribute so %*% treats them as plain vectors.
gebvs <- geno_aligned %*% as.vector(result$u) + as.vector(result$beta)
head(gebvs)

hist(gebvs,
  main = "Distribution of GEBVs (simulated data) - Grain Yield",
  xlab = "GEBV (t/ha)", col = "steelblue"
)
```

### Alternative GS Packages

The four packages solve different problems, and the choice usually follows
from what you believe about the trait rather than from performance.

**rrBLUP** assumes many markers of small effect. For a polygenic trait such
as maize grain yield that assumption is close enough to true that rrBLUP is
hard to beat, and it is the fastest of the four.

**BGLR** lets you change that assumption. Its `BRR` model is the Bayesian
equivalent of what rrBLUP fits, but `BayesB` and `BayesC` place variable
selection priors on marker effects, allowing a few markers to carry large
effects while most carry none. For a trait with known major genes, such as
resistance conferred by a single locus, those priors can outperform ridge
substantially. For yield they usually do not. BGLR also accepts multiple
terms in its `ETA` list, so you can fit markers alongside pedigree or
environmental covariates in one model.

**sommer** is the choice when the model needs more than one random effect.
Multi-trait analysis, explicit genotype by environment terms, and custom
covariance structures all belong here, and it accepts a relationship matrix
through `Gu` rather than a raw marker matrix.

**AGHmatrix** is not a fitting engine. It builds relationship matrices,
genomic from markers, pedigree-based from a known pedigree, or the combined
H matrix when only some individuals are genotyped, and hands them to
something else. Its polyploid-aware methods matter for crops such as potato
and sugarcane where the diploid assumptions elsewhere in this article do not
hold.

```{r gs-other}
# ---- BGLR (Bayesian regression) ----
library(BGLR)

bglr_dir <- tempfile("bglr_")
dir.create(bglr_dir)

fm_bglr <- BGLR(
  y = pheno_aligned$Grain_Yield_t_ha,
  ETA = list(MRK = list(X = geno_aligned, model = "BRR")),
  nIter = 1200, burnIn = 200,
  saveAt = file.path(bglr_dir, "bglr_"),
  verbose = FALSE
)
gebvs_bglr <- geno_aligned %*% fm_bglr$ETA$MRK$b
head(gebvs_bglr)

# ---- sommer (mixed model with G matrix) ----
library(sommer)
library(AGHmatrix)

# Build genomic relationship matrix
G <- Gmatrix(geno_aligned, method = "VanRaden")

fm_sommer <- mmer(
  Grain_Yield_t_ha ~ 1,
  random  = ~ vsr(germplasmName, Gu = G),
  rcov    = ~units,
  data    = pheno_aligned,
  verbose = FALSE
)
gebvs_sommer <- randef(fm_sommer)$`u:germplasmName`
head(gebvs_sommer)
```

### Cross-Validation

Fitting the model to all the data and correlating its fitted values with the
observed phenotypes would tell you almost nothing. The marker effects were
estimated from those very phenotypes, so the correlation reflects how well
the model can reproduce data it has already seen. With more markers than
individuals, which is the normal situation in genomic selection, that
correlation can approach one while the model predicts new individuals no
better than chance.

The question that matters for breeding is different: how well does the model
rank candidates it has never been phenotyped on, since the entire point is to
select before phenotyping. Cross-validation answers that directly. The
individuals are split into five folds, the model is refitted five times
holding out one fold each time, and accuracy is the correlation between
predicted and observed values in the held-out fold. Reporting the standard
deviation across folds matters too, since a mean accuracy resting on wildly
variable folds is not a stable estimate.

Two things help interpret the number. Prediction accuracy is bounded above by
the square root of heritability, because the phenotype you are validating
against is itself an imperfect measure of genetic merit. With `h2` set to 0.5
here the ceiling is about 0.71, so the accuracy reported below is a
substantial fraction of what is theoretically attainable. And random fold
assignment implicitly assumes the individuals you want to predict are related
to the ones you trained on, which is true within a breeding population but
optimistic if the real question is predicting a new family or an unrelated
germplasm source. Where that is the question, splitting by family rather than
at random gives a more honest and considerably lower estimate.

```{r cv}
set.seed(44)
n <- nrow(pheno_aligned)
folds <- sample(rep(1:5, length.out = n))
cors <- numeric(5)

for (k in 1:5) {
  train <- folds != k
  test <- folds == k

  fit <- mixed.solve(
    y = pheno_aligned$Grain_Yield_t_ha[train],
    Z = geno_aligned[train, ]
  )

  pred <- geno_aligned[test, ] %*% as.vector(fit$u) + as.vector(fit$beta)
  cors[k] <- cor(as.numeric(pred), pheno_aligned$Grain_Yield_t_ha[test])
}

cat(sprintf(
  "5-fold CV prediction accuracy (simulated data): %.3f +/- %.3f\n",
  mean(cors), sd(cors)
))
```

Prediction accuracy on simulated data is not a stand-in for accuracy on any
particular real breeding program. Its purpose here is only to demonstrate
that the pipeline (fetch -> align -> impute -> model -> validate) runs end
to end.

### A note on realism

This simulation is more realistic than the independent-marker version used
in earlier drafts of this article, but it is still a simulation, and it is
worth being precise about what it does and does not capture.

What it captures: the founder haplotypes come from a coalescent simulation
(`runMacs()`), so nearby markers are correlated the way real linked markers
are - the LD-decay check above shows this directly (adjacent-marker r²
well above random-pair r²). The population also has real family structure
from one round of random crossing among the founders, and a genetic
architecture where `n_qtl` causal sites drive the trait but never appear on
the `n_snps`-marker SNP chip the model actually sees - the chip tags the
causal variants through LD rather than containing them, which is the
realistic case: no real genotyping platform hands you the true causal
variant, only markers correlated with it.

On reproducibility: `runMacs()`'s founder haplotypes are **not** fixed by
this article's seed - AlphaSimR doesn't expose R-level seed control for
that coalescent step, so two calls to it produce different founders even
with the same `set.seed()` (see the `eval = FALSE` chunk above). Calling
it at *build* time would therefore make the cross-validation accuracy
below change on every rebuild. Instead, `runMacs()` was run once and its
result frozen to `founderPop.rds`, which this article loads instead of
regenerating - every run of this article, including the exact
cross-validation number below, is now identical. Regenerating
`founderPop.rds` (by running the `eval = FALSE` chunk again) would shift
that number to a new, equally-arbitrary value.

What it still does not capture: a single chromosome and a single
generation of random mating is a much simpler pedigree than a real
breeding program's history of selection, multiple founders per era, and
population structure across subpopulations or breeding cycles. The
heritability (`h2`) and QTL count are chosen, not estimated from any real
trait. The accuracy is realistic in magnitude for a small, single-chromosome
training population, but it describes this simulation, not any particular
real breeding program.

## References

- Selby, P., Abbeloos, R., Backlund, J.E., et al., & The BrAPI Consortium
  (2019). BrAPI — an application programming interface for plant breeding
  applications. *Bioinformatics*, 35(20), 4147–4155.
  <https://doi.org/10.1093/bioinformatics/btz190>
- Meuwissen, T.H.E., Hayes, B.J., & Goddard, M.E. (2001). Prediction of
  total genetic value using genome-wide dense marker maps. *Genetics*,
  157(4), 1819–1829. <https://doi.org/10.1093/genetics/157.4.1819>
- Endelman, J.B. (2011). Ridge regression and other kernels for genomic
  selection with R package rrBLUP. *The Plant Genome*, 4, 250–255.
- Pérez, P., & de los Campos, G. (2014). Genome-wide regression and
  prediction with the BGLR statistical package. *Genetics*, 198(2),
  483–495.
- Covarrubias-Pazaran, G. (2016). Genome assisted prediction of
  quantitative traits using the R package sommer. *PLoS ONE*, 11, 1–15.
