---
title: "Multi-Environment Trial (MET) Analysis with brapiR2"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
con <- brapiR2::brapi_connection("https://test-server.brapi.org")
server_up <- isTRUE(tryCatch(
  brapiR2::brapi_ping(con),
  error = function(e) FALSE
))
```

```{r server-down-notice, eval = !server_up, echo = FALSE, results = "asis"}
cat(
  "> **Note:** the public BrAPI test server",
  "(`https://test-server.brapi.org`) was unreachable when this article was",
  "built, so Part 1's live output below was skipped. Part 3 (simulated",
  "data) is unaffected and still ran."
)
```

## Overview

Multi-environment trial (MET) analysis is a core workflow of a plant
breeding program: gather phenotypic observations from many locations and
years, then decompose genotype x environment (GxE) interaction to identify
stable, high-performing varieties.

This article has three parts: **Part 1** fetches real study, location, and
phenotype data from the public BrAPI test server and shows exactly what
brapiR2 returns. **Part 2** shows the authenticated, parallel-fetch pattern
you'd use against your own server, without running it. **Part 3** runs the
actual GxE analysis — AMMI/GGE biplots (**metan**) and mixed-model stability
(**lme4**) — on a simulated trial, because Part 1's real data is nowhere
near large enough for either method.

## Part 1: Live Data from the Public Test Server

The three calls below return the pieces a multi-environment analysis is
assembled from. **Studies** are the environments themselves: one trial grown
at one site in one season. **Locations** carry the site metadata, latitude,
longitude, altitude and so on, that lets you group environments sensibly or
relate performance to conditions. **Seasons** identify the planting cycle,
which for maize may be a single year in temperate programs or distinct long
and short rains in East Africa.

The distinction between a location and an environment is worth being precise
about, because they are routinely conflated. An environment is a location and
season combination. The same site in two seasons contributes two
environments, and often two quite different ones, since rainfall and
temperature in a given year can matter more for yield than the site itself.
This is why a study, not a location, is the unit an AMMI or GGE analysis
treats as an environment.

```{r explore-live, eval = server_up}
library(brapiR2)
library(dplyr)

studies <- brapi_studies(con)
studies

locations <- brapi_locations(con)
locations

seasons <- brapi_seasons(con)
seasons
```

```{r pheno-live, eval = server_up}
# brapi_study_data() falls back to an unfiltered fetch + client-side filter
# when a server doesn't honour the studyDbId filter on /observations, so it
# finds data that a direct brapi_observations(con, studyDbId = ...) call
# would miss on this server. It still returns an honest empty tibble (with
# a warning) for studies that genuinely have no observations.
met_raw <- lapply(studies$studyDbId, brapi_study_data, con = con) |>
  bind_rows()
met_raw
```

Only one of these three studies (`study1`) actually has recorded
observations on the public server today - two germplasm, one trait, as the
survey at
[`dev/explore-test-server.R`](https://github.com/josh45-source/brapiR2/blob/main/dev/explore-test-server.R)
found. The other two studies come back empty. AMMI and GGE decompositions
need many genotypes scored across several environments to estimate
genotype, environment, and interaction effects separately; a mixed model
with a `germplasmName:env` random term needs enough genotype-by-environment
cells to avoid the fit being singular. Neither is possible with this data,
so Part 3 below simulates a properly-sized trial instead.

## Part 2: Authenticated, Parallel Access to Your Own Server (illustrative)

Production BrAPI servers require credentials the public test server
doesn't need, and `brapi_fetch_parallel()` is most useful across many
studies at once — this shows that pattern without running it here.

```{r setup-illustrative, eval = FALSE}
con_priv <- brapi_connection("https://my-breedbase.org")
con_priv <- brapi_login(con_priv, "username", "password")

# Cache responses to speed up re-runs and protect server bandwidth
con_priv <- brapi_cache_enable(con_priv, ttl = 7200)
```

```{r discover-illustrative, eval = FALSE}
# Discover all studies in the target trial
trials_priv <- brapi_trials(con_priv, programDbId = "wheat_program_01")
trials_priv

# Focus on the current MET
studies_priv <- brapi_studies(con_priv, trialDbId = "trial_2022")
studies_priv

study_ids_priv <- studies_priv$studyDbId
```

`brapi_fetch_parallel()` uses the **furrr** package to run one worker per
study concurrently, against whatever `future` plan is already active. It
does not set a backend itself - per the future package's best-practices
vignette, that choice belongs to the caller, not to the package, so you
set the plan yourself before calling it:

```{r parallel-illustrative, eval = FALSE}
library(future)

future::plan(future::multisession, workers = 2)
met_priv <- brapi_fetch_parallel(con_priv, brapi_study_data, study_ids_priv)
glimpse(met_priv)
future::plan(future::sequential) # shut the workers back down when done
```

## Part 3: GxE Analysis on a Simulated Trial

The design below is a 60 genotype by 6 environment trial with 2 replicates,
giving 720 plots. That is a realistic size for a regional maize yield trial
at the advanced testing stage, and it is roughly the smallest design from
which AMMI and GGE can estimate interaction patterns with any confidence.

The relative sizes of the variance components are chosen to match how yield
data actually behaves rather than to make the analysis look tidy. The
environment effect is the largest of the three systematic terms (environment,
genotype, and genotype-by-environment), consistent with site to site
variation typically being the largest systematic source of variation in a
multi-environment trial - larger than the spread between varieties within a
site, though not so dominant that it drowns out the genotype and
genotype-by-environment signal this analysis exists to recover. The genotype
by environment variance is set larger than the genotype variance, which is
the usual finding in maize METs and is precisely why this analysis exists at
all. If varieties ranked the same way everywhere, one trial would answer the
question and there would be nothing to decompose. The residual is larger
still, reflecting plot to plot variation from soil heterogeneity, stand
establishment and measurement error, which is why replication is not
optional.

```{r simulate}
library(dplyr)
library(tidyr)

set.seed(123)

n_gen <- 60L
n_env <- 6L
n_rep <- 2L

genotypes    <- sprintf("Variety_%03d", seq_len(n_gen))
environments <- sprintf("Env_%d", seq_len(n_env))

overall_mean <- 5 # tonnes per hectare

# SIMULATED variance components. Real yield heritabilities are usually
# around 0.3-0.6; sd_g/sd_ge/sd_e below are chosen to land broad-sense H2
# near 0.5, not near 1 (see Part 3's mixed model below for the fitted value).
sd_g  <- 0.35
sd_ge <- 0.55
sd_e  <- 1.00

gen_effect <- setNames(rnorm(n_gen, sd = sd_g), genotypes)
env_effect <- setNames(rnorm(n_env, sd = 0.60), environments)
ge_effect <- outer(genotypes, environments, function(g, e) {
  rnorm(length(g), sd = sd_ge)
})
dimnames(ge_effect) <- list(genotypes, environments)

met <- expand_grid(
  germplasmName = genotypes,
  env = environments,
  rep = seq_len(n_rep)
) |>
  mutate(
    observationUnitDbId = sprintf("plot_%04d", row_number()),
    Grain_Yield_t_ha = overall_mean +
      gen_effect[germplasmName] +
      env_effect[env] +
      ge_effect[cbind(germplasmName, env)] +
      rnorm(n(), sd = sd_e) # SIMULATED residual (plot-to-plot) noise
  )

nrow(met)
head(met)
```

### Means Matrix for GxE Analysis

Many GxE methods (AMMI, GGE biplot) require a two-way table of genotype
means per environment.

```{r means-matrix}
# Compute adjusted means (plot-level mean per genotype x environment)
means <- met |>
  group_by(germplasmName, env) |>
  summarise(yield = mean(Grain_Yield_t_ha, na.rm = TRUE), .groups = "drop")

# Wide format: rows = genotypes, columns = environments
means_wide <- means |>
  pivot_wider(names_from = env, values_from = yield)

means_wide
```

### AMMI and GGE Biplot with metan

Both methods start from the same observation: the two-way table of genotype
by environment means contains a pattern that main effects alone cannot
describe. They differ in what they do about it.

**AMMI** removes the additive part first, fitting genotype and environment
main effects by analysis of variance, then applies principal components to
what remains, which is the interaction. The first interaction principal
component, IPCA1, captures the dominant pattern in how varieties respond
differently across sites. In the AMMI1 biplot below, the horizontal axis is
mean yield and the vertical axis is the IPCA1 score. A genotype plotted near
zero on the vertical axis performs consistently relative to the others across
environments; one plotted far from zero has yield that depends strongly on
where it is grown. The point far to the right on mean yield and close to the
horizontal axis, high mean and near zero interaction, is the broadly
adapted variety.

**GGE** keeps the genotype main effect together with the interaction and
discards the environment main effect. The reasoning is that the environment
effect, however large, is useless for choosing a variety: knowing that one
site yields more than another tells you nothing about which line to release.
What matters for selection is genotype and its interaction, hence G plus GE.
The which-won-where view draws a polygon through the most extreme genotypes
and divides it into sectors with perpendicular rays. The genotype at the
vertex of each sector is the best performer for the environments falling in
it. When all environments land in a single sector, one variety wins
everywhere and the target region behaves as a single mega-environment. When
they split across sectors, the region does not, and that is an argument for
breeding for specific adaptation rather than for a single broadly adapted
release.

```{r metan-ammi}
library(metan)

# AMMI model: decompose GxE with additive main effects + multiplicative
# interaction
ammi_model <- performs_ammi(
  .data = met,
  env = env,
  gen = germplasmName,
  rep = rep, # replicate number, not the plot ID
  resp = Grain_Yield_t_ha,
  verbose = FALSE
)

# AMMI stability value (ASV) - lower = more stable across environments
ammi_stable <- ammi_indexes(ammi_model)
ammi_stable$Grain_Yield_t_ha |>
  arrange(ASV) |>
  head(10)

# AMMI1 biplot: IPCA1 scores vs mean yield
plot(ammi_model, type = "AMMI1")
```

```{r metan-gge}
# GGE biplot
gge_model <- gge(
  .data = met,
  env   = env,
  gen   = germplasmName,
  rep   = rep,
  resp  = Grain_Yield_t_ha
)

plot(gge_model, type = 3) # "which-won-where" view
```

### Mixed Model Stability with lme4

For unbalanced data or when you need BLUPs rather than means, fit a mixed
model with genotype x environment interaction as a random effect.

```{r lme4}
library(lme4)

# Basic mixed model: GxE interaction as random
fm <- lmer(
  Grain_Yield_t_ha ~ env + (1 | germplasmName) + (1 | germplasmName:env),
  data = met
)
summary(fm)

# Extract BLUPs for genotypic effects
blups <- ranef(fm)$germplasmName
blups <- tibble(
  germplasmName = rownames(blups),
  BLUP          = blups[["(Intercept)"]]
) |>
  arrange(desc(BLUP))

head(blups, 10)

# Variance component partitioning
VarCorr(fm)

# H^2 broad-sense heritability (across-environment), from the FITTED
# variance components (not the true simulated ones above - estimation noise
# means these won't match exactly).
var_g <- as.numeric(VarCorr(fm)$germplasmName)
var_ge <- as.numeric(VarCorr(fm)$`germplasmName:env`)
var_e <- sigma(fm)^2
n_env_obs <- length(unique(met$env))

H2 <- var_g / (var_g + var_ge / n_env_obs + var_e / (n_env_obs * n_rep))
cat(sprintf("Broad-sense H^2 (across environments): %.3f\n", H2))
```

With `sd_g = 0.35`, `sd_ge = 0.55`, `sd_e = 1.00` the true (simulated) H2 is
about 0.48; the model fitted above typically recovers something in the same
neighborhood, comfortably inside the ~0.3-0.6 range real yield
heritabilities fall in — not the unrealistically high value (0.9) that an
earlier, less careful choice of variance components produced.

### Final Selection List

The weights below, 0.6 on yield and 0.4 on stability, are a choice rather
than a standard, and they should reflect what your program is actually
selecting for.

The tradeoff behind them is real. The highest yielding line is frequently not
the most stable: it wins substantially at favourable sites and falls away
under stress, which shows up as a high mean with a large ASV. A stable line
yields moderately more or less everywhere. Which of the two a breeder should
prefer depends on the target population of environments and on who is
growing the crop. For maize under irrigation with reliable inputs, where the
conditions the variety will meet are close to the favourable end of the
trials, weighting yield heavily is defensible. For smallholder production
under variable rainfall, where a failed season carries consequences a good
season does not offset, stability deserves more weight than yield, and a
0.4/0.6 split in the other direction would be the more sensible starting
point.

One limitation worth noting: both components are rank-normalised before
combining, which makes them comparable but discards magnitude. A line
marginally ahead on yield and one far ahead contribute the same rank
difference. If the size of the yield advantage matters to the decision, scale
the components rather than ranking them.

```{r selection}
# Combine BLUP ranking with stability
stability <- ammi_stable$Grain_Yield_t_ha |>
  select(GEN, mean_yield = Y, ASV) |>
  rename(germplasmName = GEN)

selection <- blups |>
  left_join(stability, by = "germplasmName") |>
  mutate(
    # Normalise both metrics (0 = worst, 1 = best)
    rank_yield = rank(BLUP) / n(),
    rank_stab  = 1 - rank(ASV) / n(), # lower ASV = more stable = better
    score      = 0.6 * rank_yield + 0.4 * rank_stab
  ) |>
  arrange(desc(score))

# Top 10 selections: high yield + high stability
head(selection, 10)
```

As with the [genomic selection article](genomic-selection-pipeline.html),
the numbers in Part 3 describe a simulated trial and are only meant to
demonstrate that the pipeline (fetch -> reshape -> AMMI/GGE -> mixed model
-> selection) runs end to end; they say nothing about any real breeding
program.

## References

- Selby, P., Abbeloos, R., Backlund, J.E., et al., & The BrAPI Consortium
  (2019). BrAPI — an application programming interface for plant breeding
  applications. *Bioinformatics*, 35(20), 4147–4155.
  <https://doi.org/10.1093/bioinformatics/btz190>
- Zobel, R.W., Wright, M.J., & Gauch, H.G. (1988). Statistical analysis of
  a yield trial. *Agronomy Journal*, 80(3), 388–393.
  <https://doi.org/10.2134/agronj1988.00021962008000030002x>
- Yan, W., Hunt, L.A., Sheng, Q., & Szlavnics, Z. (2000). Cultivar
  evaluation and mega-environment investigation based on the GGE biplot.
  *Crop Science*, 40(3), 597–605.
  <https://doi.org/10.2135/cropsci2000.403597x>
- Olivoto, T., & Lúcio, A.D. (2020). metan: An R package for
  multi-environment trial analysis. *Methods in Ecology and Evolution*,
  11(6), 783–789. <https://doi.org/10.1111/2041-210X.13384>
- Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting
  linear mixed-effects models using lme4. *Journal of Statistical
  Software*, 67(1), 1–48. <https://doi.org/10.18637/jss.v067.i01>
